\chapter{Literature Review}

\section{Bayesian Logistic Regression for Binary Response Data}

Logistic regression is a standard member of the generalized linear model family for binary and binomial response data. In the generalized linear model framework, a response distribution from the exponential family is connected to a linear predictor through a link function; for binary response modelling, the logit link is commonly used to model the log-odds of the event probability as a linear function of covariates \cite{mccullagh1989generalized,hosmer2013applied}. This makes logistic regression suitable for modelling whether an observation belongs to an event class or a reference class.

For a binary response $y_i\in\{0,1\}$, logistic regression assumes
\[
y_i\mid p_i \sim \operatorname{Bernoulli}(p_i),
\qquad
\operatorname{logit}(p_i)
=
\log\left(\frac{p_i}{1-p_i}\right)
=
\mathbf{x}_i^T\boldsymbol{\beta}.
\]
The regression coefficient $\beta_j$ is interpreted on the log-odds scale, while $\exp(\beta_j)$ is interpreted as an odds ratio for a one-unit change in the corresponding predictor or relative to a reference category for categorical predictors \cite{hosmer2013applied}. When categorical predictors are included, one category is treated as the reference category, and the coefficients for the remaining categories describe log-odds contrasts relative to that baseline.

Bayesian logistic regression extends logistic regression by treating the coefficient vector $\boldsymbol{\beta}$ as an unknown random quantity and assigning it a prior distribution. The posterior distribution is then proportional to the product of the likelihood and the prior distribution,
\[
\pi(\boldsymbol{\beta}\mid \mathbf{y},\mathbf{X})
\propto
L(\mathbf{y}\mid \boldsymbol{\beta},\mathbf{X})\pi(\boldsymbol{\beta}).
\]
This posterior distribution represents uncertainty about the regression coefficients after observing the data \cite{gelman2013bayesian}. The Bayesian logistic regression literature is therefore relevant to studies that require uncertainty quantification for binary response models, especially when the posterior distribution must later be explored using simulation-based methods.

\section{Prior Specification in Bayesian Logistic Regression}

Prior specification is an important part of Bayesian regression modelling. In logistic regression, coefficient estimates may become unstable when the data are sparse, when predictors are imbalanced, or when separation occurs. Weakly informative priors are designed to regularise estimation without imposing strong prior beliefs. Gelman et al. proposed weakly informative Student-$t$ prior distributions for logistic and related regression models and recommended a Cauchy prior with centre $0$ and scale $2.5$ as a default prior choice after suitable scaling of predictors \cite{gelman2008weakly}.

The appeal of the weakly informative Cauchy prior is that it is centred at zero while retaining heavy tails. The centre at zero expresses a weak preference for smaller effects, while the heavy tails allow large effects when the likelihood provides sufficient support. This balance is especially relevant in logistic regression because likelihood-based estimates can become unstable under sparse data or separation. At the same time, the heavy-tailed nature of the Cauchy prior is not without consequences. Ghosh, Li and Mitra examined theoretical and computational issues associated with Cauchy priors in Bayesian logistic regression, particularly under separation, and showed that posterior behaviour may be problematic in some cases \cite{ghosh2015cauchy}. Thus, the literature supports Cauchy priors as a useful weakly informative default, but also motivates careful posterior computation and diagnostic checking.

\section{Markov Chain Monte Carlo and Metropolis--Hastings}

For many Bayesian models, posterior summaries cannot be obtained analytically because the posterior normalising constant is unavailable or posterior expectations require high-dimensional integration. Markov Chain Monte Carlo methods address this problem by constructing a Markov chain whose stationary distribution is the target posterior distribution. After burn-in, simulated draws from the chain can be used to approximate posterior means, posterior standard deviations, quantiles, credible intervals, and other summaries. Tierney provided a general theoretical treatment of using Markov chains to explore posterior distributions \cite{tierney1994markov}.

The Metropolis algorithm was introduced by Metropolis et al. for simulation-based computation \cite{metropolis1953equation}, and Hastings later generalized it to allow more general proposal distributions \cite{hastings1970monte}. The resulting Metropolis--Hastings algorithm is a flexible MCMC method for sampling from a target distribution known up to a proportionality constant. Chib and Greenberg provide a clear exposition of the algorithm and its practical implementation \cite{chib1995understanding}.

At iteration $t$, a candidate value $\boldsymbol{\beta}^{*}$ is generated from a proposal density $q(\boldsymbol{\beta}^{*}\mid\boldsymbol{\beta}^{(t-1)})$. The candidate is accepted with probability
\[
\alpha
=
\min\left\{
1,
\frac{
\pi(\boldsymbol{\beta}^{*})
q(\boldsymbol{\beta}^{(t-1)}\mid\boldsymbol{\beta}^{*})
}{
\pi(\boldsymbol{\beta}^{(t-1)})
q(\boldsymbol{\beta}^{*}\mid\boldsymbol{\beta}^{(t-1)})
}
\right\},
\]
where $\pi(\cdot)$ denotes the target density up to proportionality. This formulation covers both random-walk and independent proposals. In Random Walk Metropolis--Hastings, the proposal is centred at the current state, and symmetric proposal densities cancel from the acceptance ratio. In Independent Metropolis--Hastings, the proposal is independent of the current state, and the proposal-density correction remains in the ratio. This distinction is important in the MCMC literature because proposal design affects both theoretical validity and practical efficiency.

\section{Proposal Design and MCMC Accuracy}

The efficiency of Metropolis--Hastings sampling depends strongly on the proposal distribution. If the proposal variance is too small, the chain may have a high acceptance rate but move slowly through the posterior distribution. If the proposal variance is too large, many proposed values may be rejected. Therefore, proposal scaling is a central practical issue in Random Walk Metropolis--Hastings.

For multivariate random-walk proposals, Roberts, Gelman and Gilks developed optimal-scaling results that motivate proposal covariance matrices proportional to
\[
\frac{2.38^2}{d}\widehat{\boldsymbol{\Sigma}},
\]
where $d$ is the parameter dimension and $\widehat{\boldsymbol{\Sigma}}$ is an estimate of the target covariance structure \cite{roberts1997weak}. This result is frequently used as a theoretical guide for constructing multivariate Gaussian random-walk proposals, although practical implementations often still require additional tuning because finite-sample posteriors and applied models may not exactly satisfy the assumptions behind the asymptotic scaling result.

Independent Metropolis--Hastings has a different design requirement because its proposal is fixed rather than centred at the current state. A proposal distribution that is too narrow may fail to cover important posterior regions, while an overly diffuse proposal may generate many unlikely candidates. The literature therefore treats independent proposals as useful when a good global approximation to the target distribution is available, but potentially inefficient when the proposal tail behaviour or covariance structure does not match the target distribution well \cite{chib1995understanding,tierney1994markov}. This creates a natural basis for comparing independent and random-walk proposals in applied Bayesian computation.

Monte Carlo accuracy must also be assessed because MCMC samples are autocorrelated. Effective sample size measures the amount of independent-sample information represented by an autocorrelated chain. Monte Carlo standard error measures the simulation uncertainty of posterior estimates. Flegal, Haran and Jones emphasized that reporting Monte Carlo standard errors is important for assessing the numerical reliability of MCMC estimates \cite{flegal2008mcmc}. The literature on MCMC accuracy therefore supports reporting not only posterior summaries but also numerical measures of simulation error.

\section{Multi-chain Convergence Diagnostics}

Running multiple independent chains provides additional information about whether MCMC simulations have reached similar stationary behaviour. Gelman and Rubin proposed a multiple-sequence convergence diagnostic based on comparing between-chain and within-chain variation \cite{gelman1992inference}. This diagnostic is commonly expressed as the potential scale reduction factor, or $\hat{R}$, where values close to $1$ indicate that the independent chains have similar behaviour. Brooks and Gelman extended the diagnostic framework and discussed general methods for monitoring convergence in iterative simulations \cite{brooks1998general}.

Multi-chain diagnostics are useful because a single chain may appear stable while failing to explore all relevant regions of the target distribution. However, convergence diagnostics are not definitive proof of convergence. Cowles and Carlin reviewed a range of convergence diagnostic methods and emphasized that convergence assessment should not rely on a single numerical criterion alone \cite{cowles1996markov}. This motivates the combined use of numerical diagnostics, trace plots, autocorrelation plots, and substantive judgement.

More recent work by Vehtari et al. proposed rank-normalized, folded, and localized versions of $\hat{R}$, showing that traditional $\hat{R}$ can miss convergence problems in heavy-tailed or scale-varying chains \cite{vehtari2021rank}. Their work highlights that the classical Gelman--Rubin diagnostic is an important foundation but not the final word on convergence assessment. For literature-review purposes, this development is relevant because it shows the movement from basic multi-chain comparison toward more robust diagnostic procedures.

\section{Distributed Bayesian Computation}

Full-data MCMC can become computationally expensive for large datasets because each iteration may require evaluating the likelihood over all observations. Distributed Bayesian computation addresses this issue by partitioning the data, performing computation separately on subsets, and recombining the subset-level results. Huang and Gelman proposed an early subset-based approach for Bayesian computation with large datasets, using separate analyses on data partitions followed by approximate combination of the resulting inferences \cite{huang2005sampling}. This work is relevant because it identifies the main statistical challenge that later Consensus Monte Carlo methods address: local computations are easy to parallelise, but the local posterior information must be recombined in a way that approximates the full posterior.

The main attraction of distributed Bayesian computation is that subset-level computation can be performed with reduced memory use and greater parallelism. However, the statistical difficulty is that each subset posterior is generally less concentrated than the full-data posterior and may not contain enough information to recover the full posterior without a carefully designed aggregation method. The literature therefore treats distributed Bayesian computation as both a computational problem and a posterior-approximation problem.

\section{Consensus Monte Carlo}

Consensus Monte Carlo is a distributed Bayesian computation method designed to approximate a full-data posterior distribution using subset posterior simulations. Scott et al. proposed the Consensus Monte Carlo algorithm as a low-communication approach for Bayesian inference with large datasets \cite{scott2016bayes}. The key idea is to divide the data into shards, run separate Monte Carlo simulations on each shard, and then combine the resulting subposterior draws to approximate the full posterior distribution.

Suppose the full likelihood factorises over $K$ data shards as
\[
L(\mathbf{y}\mid\boldsymbol{\beta},\mathbf{X})
=
\prod_{k=1}^{K}
L(\mathbf{y}_k\mid\boldsymbol{\beta},\mathbf{X}_k).
\]
A common Consensus Monte Carlo construction defines the $k$th subposterior as
\[
\pi_k(\boldsymbol{\beta}\mid\mathbf{y}_k,\mathbf{X}_k)
\propto
L(\mathbf{y}_k\mid\boldsymbol{\beta},\mathbf{X}_k)
\pi(\boldsymbol{\beta})^{1/K}.
\]
The fractional prior contribution ensures that multiplying all subposterior kernels recovers a kernel proportional to the full posterior. This construction is central to the CMC literature because it clarifies how the prior should be distributed across subsets when the goal is to approximate the original full posterior rather than to fit unrelated local Bayesian models.

The remaining problem is how to combine subposterior samples. In approximately Gaussian settings, precision-weighted averaging is a natural strategy. If $\widehat{\boldsymbol{\Sigma}}_k$ is the estimated covariance matrix of subposterior $k$ and $\mathbf{W}_k=\widehat{\boldsymbol{\Sigma}}_k^{-1}$ is the corresponding precision matrix, then consensus draws can be formed using precision-weighted combinations of shard-level draws. Scott compared several Consensus Monte Carlo strategies and emphasized that the practical performance of CMC depends not only on generating local posterior samples but also on the aggregation rule used to combine them \cite{scott2017comparing}. This literature shows that simple averaging rules can be effective under approximate normality but may be less reliable when subposteriors are strongly non-Gaussian or poorly aligned.

\section{Evaluation of Consensus Monte Carlo Approximation}

Because Consensus Monte Carlo is an approximation to full-data posterior inference, the literature evaluates it by comparing the consensus approximation with a full-data posterior reference when such a reference is computationally feasible. Scott et al. and Scott compared CMC procedures using posterior summaries, approximation quality, and computational considerations \cite{scott2016bayes,scott2017comparing}. This reflects a broader principle in computational statistics: approximate algorithms should be assessed not only by speed but also by their ability to reproduce important posterior quantities.

Posterior mean error, distributional agreement, numerical stability, convergence diagnostics, effective sample size, Monte Carlo standard error, and runtime are all relevant evaluation dimensions in simulation-based Bayesian computation. Flegal, Haran and Jones argue that Monte Carlo uncertainty should be reported when simulation output is used for inference \cite{flegal2008mcmc}. Thus, evaluating a CMC approximation requires considering both approximation error relative to a reference posterior and the simulation error present in the Monte Carlo samples themselves.

The reviewed literature leaves a practical gap for applied distributed Bayesian logistic regression. Existing work establishes Bayesian logistic regression, Metropolis--Hastings sampling, proposal scaling, multi-chain diagnostics, and Consensus Monte Carlo as separate methodological components. However, the interaction between sampler choice, subposterior construction, and consensus aggregation still requires empirical evaluation in applied settings. This gap motivates the methodological design described in the next chapter.
