\chapter{Introduction}

\section{Background Study}

Bayesian inference provides a probability-based framework for learning about unknown quantities from observed data. Let $\boldsymbol{\theta}$ denote a general parameter vector and let $\mathbf{y}$ denote the observed data. Prior information about $\boldsymbol{\theta}$ is represented by the prior distribution $\pi(\boldsymbol{\theta})$, while information contributed by the data is represented by the likelihood function $L(\mathbf{y}\mid \boldsymbol{\theta})$. The posterior distribution combines these two components and is written as
\[
p(\boldsymbol{\theta}\mid \mathbf{y})
\propto
L(\mathbf{y}\mid \boldsymbol{\theta})\pi(\boldsymbol{\theta}).
\]
The posterior distribution is the central object in Bayesian inference because it describes uncertainty about $\boldsymbol{\theta}$ after observing $\mathbf{y}$. Posterior summaries such as posterior means, posterior medians, posterior standard deviations, and credible intervals are commonly used to report this uncertainty \cite{gelman2013bayesian}.

A practical weakness of Bayesian inference is that the posterior distribution is often difficult to evaluate or summarise analytically. In many realistic models, the normalising constant of the posterior distribution is unavailable in closed form, and posterior expectations require integration over a possibly high-dimensional parameter space. As a result, direct analytical calculation of posterior summaries may be infeasible, and computational approximation becomes necessary \cite{gelman2013bayesian,tierney1994markov}.

Markov Chain Monte Carlo (MCMC) methods provide a class of simulation methods for approximating expectations and posterior summaries by constructing a Markov chain whose stationary distribution is the target posterior distribution. After an adequate simulation is obtained, the generated dependent samples can be used to approximate posterior means, posterior variances, quantiles, and credible intervals. Examples of MCMC methods include the Metropolis algorithm, the Metropolis--Hastings algorithm, Gibbs sampling, Hamiltonian Monte Carlo, slice sampling, and pseudo-marginal Metropolis--Hastings \cite{metropolis1953equation,hastings1970monte,tierney1994markov}.

The Metropolis--Hastings algorithm is a general MCMC framework for sampling from a target distribution known up to a proportionality constant. It is based on a proposal distribution and an accept--reject rule. At each iteration, the algorithm proposes a candidate value for the parameter vector and accepts or rejects it according to an acceptance probability determined by the target density and the proposal distribution. Because Bayesian posterior densities are often available only up to proportionality, Metropolis--Hastings is widely used in Bayesian computation \cite{metropolis1953equation,hastings1970monte,chib1995understanding}.

This thesis focuses on two Metropolis--Hastings variants: Random Walk Metropolis--Hastings and Independent Metropolis--Hastings. In Random Walk Metropolis--Hastings, a proposed value is generated by perturbing the current value. If the proposed value is accepted, it becomes the current value for the next iteration; if it is rejected, the chain remains at the previous value. In Independent Metropolis--Hastings, proposals are generated from a fixed proposal distribution that does not depend on the current state. Since the proposal is not a symmetric random walk, a proposal-density correction is needed in the acceptance probability \cite{hastings1970monte,chib1995understanding,roberts1997weak}.

Although these Metropolis--Hastings samplers are widely applicable, each iteration requires evaluating the likelihood, and this cost grows with the number of observations. As the dataset becomes large, full-data sampling can therefore become computationally demanding, which motivates the distributed methods introduced in the following sections.

\section{Problem Statement}

Although MCMC methods are flexible, full-data Metropolis--Hastings sampling can become computationally demanding when the dataset is large. In a full-data Metropolis--Hastings algorithm, evaluating the acceptance probability typically requires repeated evaluation of the likelihood over the full dataset. As the number of observations increases, the cost of each iteration increases, and long Markov chains may require substantial computation. This computational burden motivates methods that reduce repeated full-data likelihood calculations while retaining a Bayesian posterior interpretation \cite{huang2005sampling}.

Distributed Bayesian computation addresses this issue by dividing a large dataset into smaller parts and carrying out posterior simulation or approximation at the shard level. Such approaches are motivated by settings in which the full dataset is too large to process efficiently as a single object, or where separate computations can be organised across data partitions. However, distributed Bayesian computation introduces an additional statistical problem: the shard-level posterior information must be combined in a way that approximates the full posterior distribution \cite{scott2016bayes,scott2017comparing}.

Consensus Monte Carlo is one method for distributed Bayesian computation. In the Bayesian logistic regression setting of this thesis, the parameter of interest is the coefficient vector $\boldsymbol{\beta}$, which is the logistic-regression instance of the general parameter vector $\boldsymbol{\theta}$ introduced earlier, and the full binary response vector is denoted by $\mathbf{y}$. The full dataset is divided into $K$ shards. Let $\mathbf{y}_m$ denote the response data in shard $m$, where $m=1,\ldots,K$. If the likelihood factorises across shards, then
\[
L(\mathbf{y}\mid \boldsymbol{\beta})
=
\prod_{m=1}^{K}
L(\mathbf{y}_m\mid \boldsymbol{\beta}).
\]
Consensus Monte Carlo constructs a shard-level subposterior for each shard, simulates from these subposteriors, and combines the resulting samples to approximate the full posterior distribution \cite{scott2016bayes,scott2017comparing}. A standard shard-level subposterior construction is
\[
p_m(\boldsymbol{\beta}\mid \mathbf{y}_m)
\propto
L(\mathbf{y}_m\mid \boldsymbol{\beta})\pi(\boldsymbol{\beta})^{1/K}.
\]
The factor $\pi(\boldsymbol{\beta})^{1/K}$ splits the prior contribution across the $K$ shards. This construction gives
\[
\prod_{m=1}^{K}
p_m(\boldsymbol{\beta}\mid \mathbf{y}_m)
\propto
L(\mathbf{y}\mid \boldsymbol{\beta})\pi(\boldsymbol{\beta}),
\]
so the product of the shard-level subposterior kernels is proportional to the full posterior kernel.

Several posterior combination strategies may be used in Consensus Monte Carlo, including simple averaging, scalar weighted averaging, precision-weighted averaging, and density-based product approximations. The choice of combination method matters because the consensus sample is only an approximation to the full-data posterior sample. This thesis focuses on precision-weighted Consensus Monte Carlo, where subposterior sample vectors are combined using estimated shard precision matrices so that shards with smaller posterior variance receive greater weight in the consensus combination \cite{scott2017comparing}.

The main problem considered in this thesis is how full-data Metropolis--Hastings posterior summaries compare with posterior summaries obtained from a precision-weighted Consensus Monte Carlo approximation. The comparison must be made empirically rather than assumed, because the quality of the consensus approximation can depend on shard construction, shard-level sampler behaviour, and the posterior combination rule.

Although Consensus Monte Carlo and its precision-weighted combination rule are well established methodologically, existing work has placed less emphasis on a controlled, like-for-like empirical comparison in which the same Metropolis--Hastings proposals are used for both the full-data sampler and its consensus counterpart on a single real application. This thesis addresses that gap by comparing full-data Random Walk and Independent Metropolis--Hastings directly against their precision-weighted Consensus Monte Carlo approximations under matched proposal and tuning settings. The intended contribution is therefore an empirical comparison study of approximation quality rather than a new algorithm or a claim of methodological novelty.

Road accident severity provides a suitable applied setting for this comparison. National road safety records typically contain a large number of individual accident observations, so the full-data likelihood is expensive to evaluate repeatedly and the data are naturally suited to partitioning across shards. The outcome of interest is binary, distinguishing serious/fatal accidents from slight accidents, which matches the Bernoulli likelihood structure assumed by logistic regression. Serious/fatal accidents are also far less frequent than slight accidents, so the response exhibits substantial class imbalance, and any data partition must retain enough events in each shard for stable shard-level estimation. The predictors are categorical descriptions of the road environment, which yield a moderate-dimensional coefficient vector defined through contrasts relative to reference categories. Taken together, these features---large sample size, a binary and imbalanced outcome, and a categorical predictor structure---make road accident severity a useful stress test for distributed Bayesian computation, because they probe whether the consensus approximation remains accurate when individual shards are less informative than the full dataset.

This problem is evaluated using Bayesian logistic regression for serious/fatal accident severity. The model uses categorical road type and categorical speed limit as predictors, with independent Cauchy priors for the regression coefficients. The empirical analysis compares full-data Random Walk Metropolis--Hastings and full-data Independent Metropolis--Hastings with their corresponding Consensus Monte Carlo approximations using $K=10$ shards. The CMC procedures are implemented using shard-level Metropolis--Hastings simulations and parallel worker execution, but any runtime comparison must be interpreted as specific to this implementation, computing environment, and exported results rather than as a general claim of universal speedup.

\section{Research Objectives}

The objectives of this project are as follows:

\begin{enumerate}
\item To study Bayesian inference and Metropolis--Hastings sampling as computational tools for posterior approximation.

\item To formulate a Bayesian logistic regression model for serious/fatal accident severity using categorical road type and speed limit as predictors, with independent Cauchy priors for the regression coefficients.

\item To implement full-data Random Walk Metropolis--Hastings and full-data Independent Metropolis--Hastings sampling for Bayesian posterior estimation.

\item To implement the Consensus Monte Carlo structure by dividing the data into $K=10$ shards and running shard-level Random Walk Metropolis--Hastings and Independent Metropolis--Hastings sampling.

\item To evaluate the accuracy of the precision-weighted consensus posterior approximation by comparing it with the corresponding full-data posterior summaries.

\item To compare full-data and consensus posterior summaries, credible intervals, diagnostic behaviour, acceptance rates, and recorded runtime values.
\end{enumerate}

\section{Research Framework}

The research framework begins with data preparation. The empirical data are arranged into a form suitable for Bayesian logistic regression, with the detailed preprocessing steps described in Chapter 3. This stage constructs the binary response vector $\mathbf{y}$, where serious/fatal accident severity is coded as the event of interest, and forms the predictor matrix $X$ from the selected categorical predictors.

The second stage is model specification. A Bayesian logistic regression model is defined by specifying the Bernoulli likelihood for the binary response and the prior distribution for the coefficient vector $\boldsymbol{\beta}$. The resulting posterior distribution $p(\boldsymbol{\beta}\mid \mathbf{y})$ is the target distribution for the full-data analysis.

The third stage is full-data Metropolis--Hastings sampling. Full-data Random Walk Metropolis--Hastings and full-data Independent Metropolis--Hastings are used to generate posterior samples based on the complete dataset. The retained post-burn-in samples are used to obtain full-data posterior summaries and provide the reference point for comparison.

The fourth stage is data partitioning. The full dataset is divided into $K=10$ shards, denoted by $\mathbf{y}_1,\ldots,\mathbf{y}_K$. The shard allocation is stratified by the binary response class so that the response distribution is represented across the shards. Each shard defines a shard-level likelihood contribution and a corresponding subposterior distribution with the prior contribution split across shards.

The fifth stage is shard-level Metropolis--Hastings sampling. Separate Metropolis--Hastings simulations are carried out for the subposterior distributions. The CMC shard computations are run using parallel workers, while the full-data samplers are run sequentially.

The sixth stage is subposterior combination. Samples generated from the shard-level subposteriors are combined using a precision-weighted Consensus Monte Carlo rule to produce a consensus approximation to the full posterior distribution.

The final stage compares the full-data and consensus results. The comparison focuses on posterior summaries, credible intervals, diagnostic behaviour, acceptance rates, posterior mean error measures, and recorded runtime, with runtime read descriptively from the exported results as discussed in the Problem Statement.

The seven stages of the research framework are summarised in Figure \ref{fig:research_framework}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=8mm and 0mm,
  stage/.style={
    rectangle, rounded corners, draw, thick,
    align=center, text width=0.62\textwidth,
    inner sep=6pt, font=\small
  },
  flow/.style={->, thick, >=stealth}
]
\node[stage] (s1) {Data preparation};
\node[stage, below=of s1] (s2) {Model specification};
\node[stage, below=of s2] (s3) {Full-data Metropolis--Hastings sampling};
\node[stage, below=of s3] (s4) {Data partitioning into $K$ shards};
\node[stage, below=of s4] (s5) {Shard-level Metropolis--Hastings sampling};
\node[stage, below=of s5] (s6) {Subposterior combination};
\node[stage, below=of s6] (s7) {Comparison of full-data and consensus results};

\draw[flow] (s1) -- (s2);
\draw[flow] (s2) -- (s3);
\draw[flow] (s3) -- (s4);
\draw[flow] (s4) -- (s5);
\draw[flow] (s5) -- (s6);
\draw[flow] (s6) -- (s7);
\end{tikzpicture}
\caption{The seven-stage research framework, from data preparation to comparison of full-data and consensus results}
\label{fig:research_framework}
\end{figure}

\section{Thesis Organisation}

The remainder of this thesis is organised into four chapters.

Chapter 2 presents the literature review. It surveys Bayesian logistic regression for binary response data, weakly informative prior specification, Markov Chain Monte Carlo and the Metropolis--Hastings algorithm, proposal design and Monte Carlo accuracy, multi-chain convergence diagnostics, distributed Bayesian computation, and Consensus Monte Carlo together with the precision-weighted combination rule. The chapter closes by identifying the empirical gap that motivates the methodological design of this thesis.

Chapter 3 describes the methodology. It introduces the 2013 UK accident dataset and the preprocessing steps, specifies the Bayesian logistic regression model with the Cauchy prior and resulting posterior, and explains the frequentist fit used for diagnostic checks and proposal construction. It then sets out the full-data Random Walk and Independent Metropolis--Hastings samplers, the Consensus Monte Carlo procedure (data partitioning, subposterior construction, shard-level sampling, and precision-weighted combination), and the posterior summary and evaluation criteria.

Chapter 4 reports the results and analysis. It summarises the data and shard structure, presents the frequentist logistic regression output, reports the full-data and Consensus Monte Carlo posterior summaries, and compares the four Bayesian approaches in terms of posterior estimates, acceptance rates, recorded runtime, and convergence diagnostics, supported by the exported diagnostic plots.

Chapter 5 concludes the thesis. It summarises the study and its main findings, discusses the limitations of the analysis, and outlines directions for future work.
